---
title: "21th century serious weather events consequences in USA"
author: "Arturo Garcia"
date: "26/5/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

## Synopsis
Using the data from the [NOAA's National Centers for Environmental Information (NCEI)](https://www.ncdc.noaa.gov/) we analyzed which weather events have had more impact, in terms of fatalities and property damage, in USA during the 21th century (from January 2001 to November 2011).   
The different registered weather events (more than 90) were clasiffied in 7 categories: **Cold weather** (Snow, ice, hail, ...), **Rain and Floods**, **Hot weather** (Wildfires, draught, high temperatures, ...), **Coastal** (Tide, Surf, ...), **Wind** (Hurricane, tornado, ..) and **Others** (Fog, smoke, lightning, ...).   
Total and average fatalities and costs were reported for each category across USA. We repeat the analysis grouping by State.  

## Data Preparation
Libraries:
```{r}
library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)
```



We downloaded the dataset from internet.
```{r, cache =TRUE}
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2","data.bz2")
weather <- read.csv("data.bz2", stringsAsFactors = FALSE)
```
We subset the dataset to keep only the 21th centrury records:

```{r}
weather21 <- 
  mutate(weather, date = mdy_hms(BGN_DATE)) %>% 
  select(date, COUNTYNAME, STATE, EVTYPE, FATALITIES, PROPDMG, PROPDMGEXP) %>% 
  filter(date>=dmy("01012001"), FATALITIES > 0 | PROPDMG > 0)

```

```{r}

weather21 <- weather21 %>% 
  mutate(category = case_when(
          grepl("COLD|SNOW|ICE|AVALANCHE|BLIZZARD|WINTER|HAIL|FREEZ|CHILL", .$EVTYPE) ~ "Cold", 
          grepl("COASTAL|TIDE|SURF|SEAS|RIP|TSUNAMI", .$EVTYPE) ~ "Coastal",
          grepl("WIND|STORM|TORNADO|HURRICANE|MICROBURST|TROPICAL|WATERSPOUT", .$EVTYPE) ~ "Wind",
          grepl("FLOOD|RAIN|PRECIPITATION|FLD|SEICHE|WATER", .$EVTYPE) ~ "Rain and floods",
          grepl("HEAT|FIRE|DROUGHT", .$EVTYPE) ~ "Heat",
          TRUE ~ "Other"),
        fact = case_when(
          tolower(.$PROPDMGEXP) == 'h' ~  100,
          tolower(.$PROPDMGEXP) == 'k' ~ 1000,
          tolower(.$PROPDMGEXP) == 'm' ~  1e6,
          tolower(.$PROPDMGEXP) == 'b' ~  1e9,
          TRUE ~  1), 
        damage = (PROPDMG * fact)/1e9,
        region = tolower(state.name[match(weather21$STATE, state.abb)]) ) %>% 
   select(-fact)
    
    
```

```{r}
weather21 %>% 
  group_by(category) %>% 
  summarise(sum(damage), sum(FATALITIES))
```



```{r}
states <-map_data("state")

pre <- weather21 %>% 
  group_by(category, region) %>% 
  summarise(dami = sum(damage), fat = sum(FATALITIES)) %>% 
  gather(type, value, -c(1,2)) %>% 
  group_by(region, type) %>% 
  arrange(desc(value)) %>% 
  top_n(1) %>% 
  inner_join(states) %>% 
  arrange(group)

```

```{r}
ggplot(pre) + geom_polygon(aes(x=long, y=lat, group=group, fill=category), color="white", size=.1) + coord_map() + facet_grid(type~.)
```



```{r}
lista <- weather21 %>% 
  arrange(category) %>% 
  group_by(category) %>% 
  select(EVTYPE) %>% 
  unique() 
  
```

  
